---
title: "Installing WinBugs"
author: "124384 Luycer Bosire"
date: "3/15/2021"
output:
  word_document: default
  html_document: default
---

Step 1: install and load the packages

```{r, echo = TRUE}
library("R2OpenBUGS")
library("coda")
library("readr")
````

Step 2: set working directory and load the data
````{r, echo = TRUE}
setwd("D:/Msc Statistical Science/Bayesian/Data and Codes")
wd <- getwd()
osteop <- read_delim("osteop.txt", " ", escape_double = FALSE, trim_ws = TRUE)
````

Step 3: check the loaded data
````{r, echo = TRUE}
head(osteop)
summary(osteop)
````
Step 4: write the BUGS program and put it in a txt le (one possibility). To verify the program, look at it:
````{r, echo = TRUE}
file.show("osteo.model.txt")
````

Step 5: Prepare the data and collect them into the object `data'
````{r, echo = TRUE}
N <- length(osteop$TBBMC)
data <- list (tbbmc=osteop$TBBMC, bmi=osteop$BMI, N=N)
````

Step 6: Prepare the initial values for 3 chains, we show two ways:
# One way: give the values explicitly for each chain
````{r, echo = TRUE}
inits <- list(list(beta0=0.4,beta1=0.025,tau=1/0.05),
              list(beta0=0.4,beta1=0.055,tau=1/0.05),
              list(beta0=0.4,beta1=0.025,tau=1/0.12))

# Other way: generate them randomly
inits <- function()(list(beta0=rnorm(1,0.4,1), beta1=rnorm(1), tau=runif(1,0.5,1)))
````

Step 7: Collect the parameters to monitor into one vector
````{r, echo = TRUE}
parameters <- c("beta0", "beta1", "sigma2")
````

Step 8: Run the R program bugs
````{r, echo = TRUE}

osteo.sim <- bugs(data, inits, parameters, "osteo.model.txt", n.chains=3, n.iter=1500, n.burnin = 0, n.thin=1, OpenBUGS.pgm="C:/Program Files (x86)/OpenBUGS/OpenBUGS323/OpenBUGS.exe", working.directory = wd, clearWD=FALSE, debug=TRUE, DIC=TRUE)
````

Step 9: Produce general summary of obtained MCMC sampling
````{r, echo = TRUE}
print(osteo.sim)
plot(osteo.sim)
````

Step 10: Convert osteo.sim into mcmc.list for processing with CODA
````{r, echo = TRUE}
osteo.mcmc <- as.mcmc.list(osteo.sim)
````

Step 11: Produce general summary of obtained MCMC sampling
````{r, echo = TRUE}
plot(osteo.mcmc)
summary(osteo.mcmc)
````

Step 12: Specic output obtained from CODA functions
````{r, echo = TRUE}
par(mfrow=c(2,2)) # plot figures in 2x2 format if function allows
traceplot(osteo.mcmc) # trace plots
cumuplot(osteo.mcmc,ask=FALSE) # running mean plots
acfplot(osteo.mcmc) # autocorrelation function plot
autocorr(osteo.mcmc) # autocorrelation values
crosscorr.plot(osteo.mcmc) # cross-correlation output
densplot(osteo.mcmc) # density plots of the marginal posteriors
effectiveSize(osteo.mcmc) # effective size
HPDinterval(osteo.mcmc) # HPD intervals of all parameters
````

Step 13: Other functions: convergence tests
````{r, echo = TRUE}
gelman.diag(osteo.mcmc)
gelman.plot(osteo.mcmc,ask=FALSE)
geweke.diag(osteo.mcmc)
geweke.plot(osteo.mcmc,ask=FALSE)
````

Final step: Interpret the results and make conclusions
Now we can access WinBUGS or OpenBUGS. The calls are slightly different:
## Running WinBUGS
````{r, echo = TRUE}
library("R2WinBUGS")
graft.Weib <- bugs(data = dataBUGS, inits = initsWeib, parameters.to.save = parameters, model.file = "Homograft-Bayes_ParamAFT_Weib.txt", n.chains = 3, n.burnin = 10000, n.iter = 60000, n.thin = 10, bugs.directory = "C:/Programs/WinBUGS14", working.directory = NULL, codaPkg = TRUE, debug = TRUE)

## Running OpenBUGS
library("R2OpenBUGS")
graft.Weib <- bugs(data = dataBUGS, inits = initsWeib, parameters.to.save = parameters, model.file = "Homograft-Bayes_ParamAFT_Weib.txt", n.chains = 3, n.burnin = 10000, n.iter = 60000, n.thin = 10, clearWD = FALSE, debug = TRUE)
````