---
title: "Computational Analysis Cat II"
author: "124384 - Luycer Bosire"
date: "3/10/2021"
output:
  word_document: default
  html_document: default
---

***Question One***

i. We load the data into R by first setting our working directory;

```{r, echo=TRUE}
setwd("~/Other/Msc Statistical Science/Computational Analysis/Data")
dir()

mmr.data <- read.csv("mmr.csv", header = TRUE)
head(mmr.data)

```

Use the MMR point estimates for 2017 to select 10 countries with the highest MMR;

```{r, echo=TRUE}
library(dplyr)

my.data <- subset(mmr.data,year == 2017)

my.data %>%
  group_by(iso) %>%
  arrange(desc(mmr))

use.data <- my.data[1:10 ,]
use.data
```
a. sort country name alphabetically

```{r, echo = TRUE}
iso.data <- use.data[order(use.data$iso), ]
iso.data
```

b. the MMR point estimate for 2017 and the lower and upper bound

```{r, echo = TRUE}

my.data <- subset(mmr.data,year == 2017)

point.estimate.data <- my.data$mmr

#A point estimate is the mean 
point.estimate <- mean(point.estimate.data, na.rm = TRUE)
point.estimate

#Upper bound and lower bound
upper.bound <- subset(my.data,bound == "upper")
upper.bound.data <- upper.bound$mmr
upper.point.estimate <- mean(upper.bound.data)

lower.bound <- subset(my.data,bound == "lower")
lower.bound.data <- lower.bound$mmr
lower.point.estimate <- mean(lower.bound.data)

sd_hat <- sd(point.estimate.data, na.rm = TRUE)
upper_bound <- sd_hat
lower_bound <- (sd_hat*-1)

bounds <- c(lower_bound,upper_bound)
round(bounds, digits = 3)

estimate<- c(point.estimate, lower.point.estimate, upper.point.estimate, bounds)
estimate
```

c. annual rate of reduction (AAR) for 2000 - 2017

d. estimated number of maternal deaths for 2017 as well as the upper and lower bounds
```{r, echo = TRUE}

setwd("~/Other/Msc Statistical Science/Computational Analysis/Data")

matdeaths.data <- read.csv("matdeaths.csv", header = TRUE)
head(matdeaths.data)

point.estimate.data <- matdeaths.data$value

#A point estimate is the mean 
point.estimate <- mean(point.estimate.data, na.rm = TRUE)
point.estimate

#Upper bound and lower bound
upper.bound <- subset(matdeaths.data,bound == "upper")
upper.bound.data <- upper.bound$value
upper.point.estimate <- mean(upper.bound.data)

lower.bound <- subset(matdeaths.data,bound == "lower")
lower.bound.data <- lower.bound$value
lower.point.estimate <- mean(lower.bound.data)

sd_hat <- sd(point.estimate.data, na.rm = TRUE)
upper_bound <- sd_hat
lower_bound <- (sd_hat*-1)

bounds <- c(lower_bound,upper_bound)
round(bounds, digits = 3)

matdeaths.estimate<- c(point.estimate, lower.point.estimate, upper.point.estimate, bounds)
matdeaths.estimate
```
e. the share of maternal deaths as a proportion of the global total for 2017 for each of the 10 countries

Output your results in csv format
```{r, echo=TRUE}
write.csv(use.data, file = "UseData.csv")
write.csv(estimate, file = "Estimates.csv")
write.csv(matdeaths.estimate, file = "MatDeathsEstimates.csv")

dir()
```

ii. Loading this data into r;
```{r, echo = TRUE}
mytable.data <- read.csv("arsenic.csv", header = TRUE)
head(mytable.data, 3)
```

***Question Two***

i. If $X \sim Exp(1),$ then $F(x)=1 - exp(-x).$ Solving $x$ in $u=1 - exp(-x)$ gives:

$$u=1-exp(-x)\\exp(-x)=1-u\implies X=-log (U)\sim exp(1).$$

Compare the random generation between uniform transformation and the exponential distribution using 10000 random numbers
```{r, echo = TRUE}

Nsim=10^4 #number of random variables
U=runif(Nsim)
X=-log(U) #transforms of uniforms
Y=rexp(Nsim) #exponentials from R
par(mfrow=c(1,2)) #plots
hist(X,freq=F,main="Exp from Uniform")
hist(Y,freq=F,main="Exp from R")

```

The fits of both histograms to their exponential limit are not distinguishable.The representations can always be generated from deterministic uniform random variables.

ii. load the library ape and produce a pie chart
install.packages("ape")

```{r, echo = TRUE}

library(ape)
library(plotrix)

bank<-table(read.GenBank(c("X94991.1"), as.character=TRUE))
bank

colors = c('#4286f4','#bb3af2','#ed2f52','#efc023')
labels=c("A","B","C","D")
pie(bank, labels, main='GenBank', col=colors, init.angle=180, clockwise=TRUE)

colors = c('#4286f4','#bb3af2','#ed2f52','#efc023')
labels=c("A","B","C","D")
pie3D(bank, labels=labels, explode=0.1, height=0.05, main='Genbank 3D Piechart', col=colors)

```

iii. Use the file "gapminder.csv' to perform the following analysis;
```{r, echo=TRUE}
data_set<-read.csv("gapminder.csv")
str(data_set)
head(data_set)
```

a. Obtain data set for only the year 1982

```{r, echo=TRUE}
data_1982<- subset(data_set,subset = year==1982)
head(data_1982)
str(data_1982)
```
b. Obtain data set for the American countries in 1997

```{r, echo=TRUE}
data_America<-subset(data_set,continent == "Americas"& year==1997)
head(data_America)
```

c. Add a column to the data set loaded in R with the new column being gdp which is computed as gdp=popgdpPercap*

```{r, echo=TRUE}
attach(data_set)
GDP<-function(pop,gdpPercap){pop*gdpPercap}
gdp<-GDP(pop,gdpPercap)
head(gdp)

data_set_1<- cbind(data_set,gdp)
str(data_set_1)
```
