---
title: "Bayesian Cat I"
author: "124384 Luycer Bosire"
date: "2/6/2021"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Question 1: Single parameter poisson**

Consider the count of airline crashes per year over a 10-year period form 1976 to 1985:
  
    y <- c(24,25,31,31,22,21,26,20,16,22)
    n<- length(y)
    n
    bar_x<- mean(y)
    bar_x

**_a. Show that the Gamma distribution is the conjugate prior for a Poisson mean._**

Solution in the steps below; 

1. write down the likelihood $p(X|\lambda)$ for $\lambda$. 

The pdf and likelihood function of Poisson Distribution with parameter $\lambda$ is given by;

  $$ P(x| \lambda) = \frac {e^{-\lambda}\lambda^{x}}{x!} $$
  
$$ L(\lambda; \pmb x) = \frac {e^{-n\lambda}\lambda^{n\bar x}}{\prod_{i=1}^nx_i!}\propto e^{-n\lambda}\lambda^{n\bar x} $$

2. Write down the prior density for $\lambda$.

The prior density Gamma with parameter $$n, \lambda$$ is given by;

$$ P(x |\lambda)= \frac {\beta^\alpha} {\Gamma(\alpha)} {\lambda^{(\alpha-1)}} {e^ {(-\lambda\beta)}} $$

Ignoring the constant term:
$$P(\lambda)=\lambda^{\alpha-1}e^{-\beta\lambda}$$
3. Multiply them together to obtain the posterior density, and notice that it has the same form as the gamma distribution.

Using Bayes Rule

$$\text{posterior} \propto \text{likelihood} \times \text{prior},$$
That is
$$P(\lambda|x)=P(x|\lambda) \times P(\lambda)$$

Therefore:
$$P(\lambda|x)=e^{-n\lambda}\lambda^{n\bar x}\times\lambda^{\alpha-1}e^{-\beta\lambda}$$
$$P(\lambda|x)=\lambda^{n\bar x+\alpha-1}e^{-\lambda(n+\beta)}$$
Thus we can notice that the posterior takes a Gamma form:
$$Post(\lambda|x)∼Gamma(\alpha^*,\beta^*)$$
Where
$$\alpha^*=n\bar x+\alpha $$
$$\beta^*=n+\beta$$

**_b. Show graphics of the resulting posterior distribution for choices of various gamma priors_**

Confirming the code for different values of alpha and beta:

```{r}
library(shiny)
library(tidyverse)

# ui codes ----------------------------------------------------------------

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
    sliderInput("n", "Play to increase sample size", min = 1,
                max = 10, step = 1,value = 10, animate = T),
    tags$hr(),
    numericInput("beta", "Beta value", value = 5),
    tags$hr(),
    numericInput("alpha", "Alpha Value", value = 5), width = 2
    ),
    
    mainPanel(
      h3("The likelihood and posterior distributions converge as the sample size n increases"),
      br(),
plotOutput("my_plot", height = "500px")
    )
    
  )
)

# server code -------------------------------------------------------------

server <- function(input, output, session){
  


# define posterior function -----------------------------------------------
y <- c(24,25,31,31,22,21,26,20,16,22)
Lambda <- mean(y)
post <- function(n = 10,alpha = 5, beta = 1/Lambda){

lambda<-Lambda  # 1data initialization

LL <- dpois(y,lambda)  # quartiles of a binomial

# prior
Prior <- dgamma(y,shape = alpha,scale = beta)  #beta prior distribution
  
# posterior
alpha1 <- n*lambda + alpha
beta1 <- n+ beta

Postr <- dgamma(y,shape=alpha1,scale = beta1)

ggplot(data = NULL, aes(y, LL/max(LL), col = "Likelihood")) + geom_line(size = 1.0)+
  geom_line(aes(y, Prior/max(Prior), col = "Prior"), size = 1.0) + 
  geom_line(aes(y, Postr/max(Postr), col = "Posterior"), size = 1.0) +
  labs(y = "Density", x= expression(y)) + theme_minimal() +
  scale_colour_manual("", values = c("Likelihood"="purple", "Prior"="red", "Posterior"="blue")) +
  theme(legend.position = "top")

}

output$my_plot <- renderPlot({
  post(n=input$n,alpha = input$alpha, beta = input$beta)
})
  
  
}

shinyApp(ui, server)

```
**_c. Show a table of the prior, MLE and posterior estimates of the poison mean under different choices of the gamma priors in (a) above._**

1.When $\alpha=3$ and $\beta=6$
```{r}

Table_Estimates <- function(alpha,beta){
  
y <- c(24,25,31,31,22,21,26,20,16,22)
lambda <- mean(y)
MLE <- dpois(y,lambda)
Prior01 <- dgamma(y,shape = alpha,scale = beta)
Postr01 <- MLE*Prior01

Estimates <- rbind(MLE,Prior01,Postr01)

return(Estimates)

}
df <- data.frame(Table_Estimates(3,6))
df
```
**QUESTION TWO**

**_a.Differentiate between Credible Intervals and the Highest Posterior Density (HPD) in Bayesian analysis._**



A credible interval is an interval within which an unobserved parameter value falls with a particular probability. It is an interval in the domain of a posterior probability distribution or a predictive distribution.On the other hand, a highest posterior density (interval) is basically the shortest interval on a posterior density for some given confidence level.


**_b.Graphically demonstrate this for a disease prevalence given that out of 150 individuals 18 had the disease. Hint: assume Binomial distribution_**

Highest Posterior Density (HPD)
```{r}
library(binom)
HPD <- binom.bayes(x=18,n=150,type = "highest",conf.level = 0.95,tol = 1e-9)
print(HPD)
binom.bayes.densityplot(HPD)
```

Credible Interval

```{r}
library(binom)
Central <- binom.bayes(x=18,n=150,type = "central",conf.level = 0.95,tol = 1e-9)
print(Central)
binom.bayes.densityplot(Central)

```
