---
title: "CAT I Question I"
author: "124384 Luycer Bosire"
date: "2/7/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Bayesian Cat I"
author: "124384 Luycer Bosire"
date: "2/6/2021"
output: html_document
---

**Question 1: Single parameter poisson**

Consider the count of airline crashes per year over a 10-year period form 1976 to 1985:

**_a. Show that the Gamma distribution is the conjugate prior for a Poisson mean._**

Solution in the steps below; 

1. write down the likelihood $p(X|\mu)$ for $\mu$. 

The pdf and likelihood function of Poisson Distribution with parameter $\mu$ is given by;

$$ f(x, \mu) = \frac {e^{-\mu}\mu^{x}}{x!} $$
  
$$ L(\mu; \pmb x) = \frac {e^{-n\mu}\theta^{n\bar x}}{\prod_{i=1}^nx_i!}\propto e^{-n\mu}\mu^{n\bar x} $$

2. Write down the prior density for $\mu$.

The prior density Gamma with parameter $$n, \lambda$$ is given by;

$$ f(n, \lambda)= \frac {\beta^\alpha} {\Gamma(\alpha)} {\lambda^{(\alpha-1)}} {e^ {(-\lambda\beta)}} $$
3. Multiply them together to obtain the posterior density, and notice that it has the same form as the gamma distribution.

Using Bayes Theorem which we can write in words as
$$\text{posterior} \propto \text{likelihood} \times \text{prior},$$

In this case, the likelihood $p(D | \mu)$ is given by
$$p(D | \mu) \propto e^{-n\mu}\mu^{n\bar x}$$ 
```{r, echo=TRUE}

    y <- c(24,25,31,31,22,21,26,20,16,22)
    n<- length(y)
    n
    bar_x<- mean(y)
    bar_x
```

where n = 10 and mean = 23.8 so our likelihood is
$$p(D | \mu) \propto e^{-10\mu}\mu^{10*23.8}$$

Combining the prior and the likelihood we get;

$$ p(D | \mu) \propto (e^{-10\mu}\mu^{10*23.8}) ({\frac {\beta^\alpha} {\Gamma(\alpha)} {\lambda^{(\alpha-1)}} {e^ {(-\lambda\beta)}}}) \propto \lambda ^{(n+\alpha-1)} e^{-(n+\beta)\lambda} $$
At this point we again apply the "trick" of recognizing this density as the density of a Gamma distribution - specifically, the Gamma distribution with parameters 

$$(\alpha+n \bar x,\beta+n) \propto (\alpha+10*23.8,\beta+10)$$.

